The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-06-08 guides/practical_guides.Rmd
#15 Added Github actions
spelling, editing, BiB formatting

### Intro to R and Rstudio

@@

### Shell and linux systems

@@

### Intro to Python

@@

### Intro to VS Code

@@

Test the reproduction process as if you were an external researcher.

@@

>>>>

### Introduction to R and Rstudio

@@

### Introduction to Unix and Shell programming

@@

### Introduction to Python

@@

### Introduction to VS Code

@@

Continuous analysis involves automatic execution and testing of code, which is important when changes
are made to code or data.
Results generated in HTML are automatically updated.
Badges on the repository indicate the status of the codebase in terms of completing with or without
errors.

@@

* [Documentation: GitHub Actions](https://docs.github.com/en/actions)

### 2023-06-08 manuscript/5pillars.Rmd
spelling, editing, BiB formatting

Computational reproducibility is a simple premise in theory, but in practice it is difficult to
achieve.
Building upon past efforts and proposals to maximise reproducibility and rigour in
bioinformatics, we present a framework called the five pillars of reproducible computational
research.
These include (1) code version control and sharing, (2) compute environment control,
(3) literate programming, (4) documentation and (5) FAIR data sharing.
These best practices will ensure that computational research work can be
reproduced quickly and easily, long into the future.
This guide is designed for bioinformatic data analysts, but should be relevant for other
domains.

@@

A number of guides recommending enhanced computational reproducibility practices have been developed
[ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
@Wilson2017-uq ; @Akalin2018-eg ; @Gruning2018-sg; @Brito2020-ld; @Peikert2021-eq ; @Rodrigues2023-fw;
@DuPre2022-bk].
Inspired by these principles, this article presents the
*five pillars of reproducible computational research* (**Figure 1**).
There are clear parallels with the established the three pillars of open science framework,
(open data, code and papers) [@Gorgolewski2016-hv].
Here, the emphasis is on practical reproducibility with an increased focus on programming practices,
transparent reporting and the role of computational environments.
The intended audience is bioinformatics data analysts (not tool developers), however the
principles described here could equally apply to other domains of study.

@@

For a bioinformatics project to be reproducible, it needs to be enshrined by code and include all
instructions, from inspecting the raw data to generating the outputs that form the conclusions of the
study.
This automated process removes the need for any manual steps, which are not only time-consuming but
also prone to errors.
Without an end-to-end automated process, most of the other best practices described here are not
possible.
Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
which would be difficult for graphical tools like spreadsheets or webtools.
Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
somewhat overused and misused in the life sciences, leading to problems like inadvertant conversion
of gene names to dates in databases and supplementary Excel files [@Zeeberg2004-js;
@Ziemann2016-pv; @Abeysooriya2021-uz].
While webtools are valuable for data exploration, there are worries that they undermine
reproducibility for the sake of convenience [@Wijesooriya2022-ly].
Transferring data between compute platforms is also discouraged.
For example, having workflows that involve combinations of web-based and scripted tools require
data transfer steps which are inefficient and error-prone.

@@

A guiding principle of the five pillars approach is that the publishing researchers should strive
to make the reproduction process easier and faster for those who wish to.
A "master script" that coordinates the execution of individual parts of an analysis is an excellent
way to reduce the number of commands required for reproduction.
This can be coded in standard open source languages like R/R Markdown, Python/Jupyter notebook or the
Unix Shell.
Using free and open source software ensures that the research is accessible to the greatest audience,
as opposed to proprietary software like SPSS, STATA and MatLab which are cost inhibitory.
Free and open source software also enjoy relatively larger libraries of add-on packages contributed by
the scientific community.
Projects entailing big data analysis might benefit from a build/workflow automation solution like `make`,
`snakemake` [@Koster2012-sq; @Molder2021-mp], `drake` [@Michael_Landau2018-kk] or
`nextflow` [@Di_Tommaso2017-yo].
The advantage of such tools is that if the analysis were to terminate due to a hardware issue mid-way,
for example at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't need to be repeated.
After fixing the hardware issue, re-running the analysis would pick up at step 8 again,
which saves labour and compute time.

@@

Literate programming combines "chunks" of analytical code with human readable text @Knuth1984-di.
After compilation, the resulting output document contains the code together with computational results
such as figures and tables along with contextualising explanations and narratives.

@@

* Helps collaboration and project management.
Using a centralised code hosting platform such as "GitHub" or "BitBucket" can help manage contributions
from team members who may be working asynchronously in different time zones and on different servers.
It also makes the user's work searchable, which is convenient when code snippets from a project
need to be reused for a new project.

* Helps resolve issues.
These centralised platforms also enable discussions of code issues such as bugs and feature requests.
In bioinformatics teams, issues can be used to track the progress of a project through its milestones
and allow team members to raise potential issues with the code.
This also allows the scientific community to raise an issue with the software maintainers.
For example, this feature is used extensively during the peer review process for the *Journal of
Open-Source Software* @Editorial2018-kt.

@@

Although DVCSs assist with code sharing, they are not considered a long-term archiving solution, 
as recalled by the complete shutdown of the Google Code platform in 2015 
and the end of Mercurial support by the Bitbucket platform in 2020. 
The code needs to be deposited to a long-term repository.
[Zenodo](https://zenodo.org/) and [Figshare](https://figshare.com/) are often used 
to store the content of a repository but do not retain the history of the development of the source code.
[Software Heritage](https://www.softwareheritage.org/) @dicosmo2017 is the universal software archive 
which mission is to collect, preserve and share source code. Software Heritage provides 
an intrinsic persistent identifier called SWHID that allows to cite the archived source code 
in the respective journal article/preprint. The archival of the source code in Software Heritage
can be performed manually or automated by a webhook within a continuous integration.

@@

Guix is described as a "functional package manager" and has the unique benefit of giving
bit-for-bit build reproducibility and verifiability [@Vallet2022-no; @Courtes2022-rx].
This is a notable advantage, as Docker images are not verifiable nor guaranteed to build reproducibly
in the future due to link decay.
Moreover, Guix can build and output environments as Docker-compatible images which solves the
provenance problem of such environments [@GNU_Project_Volunteers2022-bt].
These features are being used in the bioinformatics sphere to create highly reproducible pipelines
that work equally well on personal as well as shared high performance computers
[@Wurmus2018-lv; @Courtes2022-rx]. 
Guix is probably one of the best solution so far for the "ten-year challenge" of code reproducibility @Perkel2020-io.
Although Guix has some exciting functionality, there remains a relative lack of step-by-step guides and
tutorials, illustrated by the complete absence of published Guix protocols in protocols.io, as compared
to 13 and 12 related to "Conda" and "Docker" respectively (as at 30th May 2023).
Eventually, it's also worth mentionning that the Guix project uses Software Heritage
as a fallback system to provide a package source code
if the download of the source code from its original location failed.

@@

Without data sharing, computational research is not reproducible nor auditable.
Lack of data sharing is one of the key reasons why research is irreproducible
[@Ioannidis2009-hv, @Archmiller2020-er ].
Data sharing is also one of the key features of "open science", which is characterised as science
that is collaborative, transparent, accessible and inclusive [@McKiernan2016-vu; @noauthor_2023-uq].
In addition to facilitating reproduction and auditing, sharing enables reuse in other contexts.
Data reuse increases efficiency, as it prevents redundant research expenditure and facilitates new
research ideas that were previously impossible [@Wilkinson2016-qq].
In terms of research rigour, inspecting raw data can uncover inadvertent errors and research
integrity problems [@Miyakawa2020-yj].
In light of this, a research article without supporting data and code is much like a press release or
advertisement where the claims made cannot be verified as true or not [@Donoho2010-yz, @DuPre2022-bk].

Although it is common to see journal articles with "Data available upon reasonable request", this is less
than ideal.
A systematic study of such data availability statements found that the data was successfully provided
in just 6.8% of requests [@Gabelica2022-ci].
In the rare cases that data is shared in accordance with the data availability statement, it poses
a burden in terms of labour to lodge and respond to such requests, and if the data is large, may incur
an additional cost for hosting/transferring these large files.

@@

Some researchers would rather keep data for their private use.
Some may keep data private due to concerns over participant privacy.
This is a well-founded concern for identifiable data, however once identifying features are removed,
it is nearly impossible to reidentify genomic data @Venkatesaramani2021-tn unless the attackers have
preexisting genomic data describing an individual or ancestry [@Schadt2012-md; @Gymrek2013-fz].
Therefore, sharing deidentified genomic data and metadata is likely to carry little risk to participant
privacy.
Still, much human genomic data and metadata resides in restricted databases due to the scope of
data sharing that was agreed to in the consent process.
For example, the consent form may only allow sharing of data to researchers investigating certain
diseases, which means a vetting process must be undertaken before sharing data with external
investigators.
In the interests of reuse and reproducibility, future studies could consider a consent form that allows
more broad sharing of deidentified data.

In order to maximise the value of shared data for reuse and reproducibility, it needs to be Findable,
Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq.
These principles are summarised in **Figure 5**.

@@

High throughput biology has already built a number of data repositories for specific data types
conforming to FAIR principles (e.g.: GEO, SRA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
However other domains lack well known specialised organised data repositories, such as evolutionary
biology, ecology, cell biology, social sciences and computing.
The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find a
repository that accepts data from these fields [@Pampel2013-ep].
If no specialised repository exists, then a general-purpose repository such as
[Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
should be considered.
Moreover, general recommendations have been proposed to enhance data FAIRness in ecology, but which
can also be applied to other disciplines [@Roche2015-aq].

@@

1. Depositing data to a specialised repository if possible, otherwise, a general-purpose repository.

@@

Once the quality of the data set has been established researchers may consider depositing it
to a data repository early, before publicisation, as it has some advantages.
Most data repositories have an optional 12-month embargo period so researchers can share publicly
at a later date once they're happy with the overall project.
The data repository acts as an additional backup to the host institutions' own in case of data loss
or calamity.
By assuming the data and metadata can be sourced from a public location, the workflow developers ensure
that the code they share won't suffer from "file not found errors" during reproduction.
This ensures code and data are linked, which has been a priority for reproducibility [@Peng2011-ia],
and a source of problems for shared code [@Samuel2022-ha].

In addition to the experimental data, reference data that is critical to a project should be archived.
For example, in the field of genomics, gene function annotation sets are relied upon for a common
procedure called "pathway analysis" @Khatri2012-ye.
These annotation sets are regularly updated, but versions are rarely reported [@Wijesooriya2022-ly]
and finding historical data versions is sometimes difficult.
In order to guarantee future reproducibility, snapshots of those reference data should be archived
and shared if the licence permits it.

@@

Documentation is the glue that binds a data science project together.
The manuscript/published article is the central artefact that outlines the research project and
links to the supporting materials.
From a reproducibility standpoint, the methods section is the most critical part.
It should be detailed enough so that other researchers can understand and replicate the
experiments/analysis and yield similar results.
Unfortunately, key details in bioinformatics data processing procedures are often omitted which limits
their reproducibility [@Simoneau2021-kp].
The "Materials Design Analysis Reporting" (MDAR) checklist for authors has been developed to assist in
comprehensive methodological reporting in the life sciences [@Mellor2022-up], and "Minimum Information
About a Bioinformatics Investigation" guidelines describe good reporting practices for computational
biology research [@Tan2010-is].

@@

As outlined above, one of the goals is to reduce the complexity of reproduction, and using a master
script is a simple way to do this.
It also makes the `README` documentation much simpler.
For example, the pathway analysis workflow we previously developed could be reproduced and inspected
with just six commands, starting from a new Ubuntu Linux install [@Ziemann2023-nm].
By minimising the complexity of reproduction and writing the instructions clearly in the `README`,
reproducibility is made possible to a larger segment of the computational research community,
especially those who are not experts in the languages used.

@@

A lack of technical computer skills in many research groups is another culprit.
Specific workshop-based training in these areas could help, but the most scalable solution would be to
provide comprehensive beginner-friendly step-by-step guides for frequently used routines.
For example, our group has just released such a protocol that walks novice users through the process
of implementing functional enrichment analysis of gene expression data with the five pillars principles
@Ziemann2023-nm.
More such protocols are needed to make extreme reproducibility mainstream.
A curated set of tutorials is provided in the Supplement.

@@

Reproducibility is an essential feature of reliable computational research yet current practices
are falling short of this goal.
Our recommendations for workflow-based research articles are summarised as:

@@

@@

>>>>

Computational reproducibility is a simple premise in theory, but in practice it is difficult to
achieve.
Building upon past efforts and proposals to maximize reproducibility and rigor in
bioinformatics, we present a framework called the five pillars of reproducible computational
research.
These include (1) code version control and sharing, (2) compute environment control,
(3) literate programming, (4) documentation and (5) FAIR data sharing.
These best practices will ensure that computational research work can be
reproduced quickly and easily, long into the future.
This guide is designed for bioinformatic data analysts, but should be relevant for other
domains.

@@

A number of guides recommending enhanced computational reproducibility practices have been developed
[ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
@Wilson2017-uq ; @Akalin2018-eg ; @Gruning2018-sg; @Brito2020-ld; @Peikert2021-eq ; @Rodrigues2023-fw;
@DuPre2022-bk].
Inspired by these principles, this article presents the
*five pillars of reproducible computational research* (**Figure 1**).
There are clear parallels with the established three pillars of the open science framework,
(open data, code and papers) [@Gorgolewski2016-hv].
Here, the emphasis is on practical reproducibility with an increased focus on programming practices,
transparent reporting and the role of computational environments.
The intended audience is bioinformatics data analysts (not tool developers), however the
principles described here could equally apply to other domains of study.

@@

For a bioinformatics project to be reproducible, it needs to be enshrined by code and include all
instructions, from inspecting the raw data to generating the outputs that form the conclusions of the
study.
This automated process removes the need for any manual steps, which are not only time-consuming but
also prone to errors.
Without an end-to-end automated process, most of the other best practices described here are not
possible.
Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
which would be difficult for graphical tools like spreadsheets or webtools.
Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
somewhat overused and misused in the life sciences, leading to problems like inadvertent conversion
of gene names to dates in databases and supplementary Excel files [@Zeeberg2004-js;
@Ziemann2016-pv; @Abeysooriya2021-uz].
While webtools are valuable for data exploration, there are worries that they undermine
reproducibility for the sake of convenience [@Wijesooriya2022-ly].
Transferring data between compute platforms is also discouraged.
For example, having workflows that involve combinations of web-based and scripted tools require
data transfer steps which are inefficient and error-prone.

@@

A guiding principle of the five pillars approach is that the publishing researchers should strive
to make the reproduction process easier and faster for those who wish to.
A "master script" that coordinates the execution of individual parts of an analysis is an excellent
way to reduce the number of commands required for reproduction.
This can be coded in standard open source languages like R/R Markdown, Python/Jupyter notebook or the
Unix Shell.
Using free and open source software ensures that the research is accessible to the greatest audience,
as opposed to proprietary software like SPSS, STATA and MatLab which are cost inhibitory.
Free and open source software also enjoy relatively larger libraries of add-on packages contributed by
the scientific community.
Projects entailing big data analysis might benefit from a build/workflow automation solution like `make`,
`snakemake` [@Koster2012-sq; @Molder2021-mp], `drake` [@Michael_Landau2018-kk] or
`nextflow` [@Di_Tommaso2017-yo].
The advantage of such tools is that if the analysis were to terminate due to a hardware issue mid-way,
for example at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't need to be repeated.
After fixing the hardware issue, re-running the analysis would pick up at step 8 again,
which saves labor and compute time.

@@

Literate programming combines "chunks" of analytical code with human readable text @Knuth1984-di.
After compilation, the resulting output document contains the code together with computational results
such as figures and tables along with contextualizing explanations and narratives.

@@

* Helps collaboration and project management.
Using a centralized code hosting platform such as "GitHub" or "BitBucket" can help manage contributions
from team members who may be working asynchronously in different time zones and on different servers.
It also makes the user's work searchable, which is convenient when code snippets from a project
need to be reused for a new project.

* Helps resolve issues.
These centralized platforms also enable discussions of code issues such as bugs and feature requests.
In bioinformatics teams, issues can be used to track the progress of a project through its milestones
and allow team members to raise potential issues with the code.
This also allows the scientific community to raise an issue with the software maintainers.
For example, this feature is used extensively during the peer review process for the *Journal of
Open-Source Software* @Editorial2018-kt.

@@

Although DVCSs assist with code sharing, they are not considered a long-term archiving solution, 
as recalled by the complete shutdown of the Google Code platform in 2015 
and the end of Mercurial support by the Bitbucket platform in 2020. 
The code needs to be deposited to a long-term repository.
[Zenodo](https://zenodo.org/) and [Figshare](https://figshare.com/) are often used 
to store the content of a repository but do not retain the history of the development of the source code.
[Software Heritage](https://www.softwareheritage.org/) @dicosmo2017 is the universal software archive 
whose mission is to collect, preserve and share source code. Software Heritage provides 
an intrinsic persistent identifier called SWHID that allows to cite the archived source code 
in the respective journal article/preprint. The archival of the source code in Software Heritage
can be performed manually or automated by a webhook within a continuous integration.

@@

Guix is described as a "functional package manager" and has the unique benefit of giving
bit-for-bit build reproducibility and verifiability [@Vallet2022-no; @Courtes2022-rx].
This is a notable advantage, as Docker images are not verifiable nor guaranteed to build reproducibly
in the future due to link decay.
Moreover, Guix can build and output environments as Docker-compatible images which solves the
provenance problem of such environments [@GNU_Project_Volunteers2022-bt].
These features are being used in the bioinformatics sphere to create highly reproducible pipelines
that work equally well on personal as well as shared high performance computers
[@Wurmus2018-lv; @Courtes2022-rx].
Guix is among the best solutions for the "ten-year challenge" of code reproducibility @Perkel2020-io.
Although Guix has some exciting functionality, there remains a relative lack of step-by-step guides and
tutorials, illustrated by the complete absence of published Guix protocols in protocols.io, as compared
to 13 and 12 related to "Conda" and "Docker" respectively (as at 30th May 2023).
Lastly, it's worth mentioning that the Guix project uses Software Heritage
as a fallback system to provide a package source code
if the download from its original location failed.

@@

Without data sharing, computational research is not reproducible nor auditable.
Lack of data sharing is one of the key reasons why research is irreproducible
[@Ioannidis2009-hv, @Archmiller2020-er ].
Data sharing is also one of the key features of "open science", which is characterized as science
that is collaborative, transparent, accessible and inclusive [@McKiernan2016-vu; @noauthor_2023-uq].
In addition to facilitating reproduction and auditing, sharing enables reuse in other contexts.
Data reuse increases efficiency, as it prevents redundant research expenditure and facilitates new
research ideas that were previously impossible [@Wilkinson2016-qq].
In terms of research rigor, inspecting raw data can uncover inadvertent errors and research
integrity problems [@Miyakawa2020-yj].
In light of this, a research article without supporting data and code is much like a press release or
advertisement where the claims made cannot be verified as true or not [@Donoho2010-yz, @DuPre2022-bk].

Although it is common to see journal articles with "Data available upon reasonable request", this is less
than ideal.
A systematic study of such data availability statements found that the data was successfully provided
in just 6.8% of requests [@Gabelica2022-ci].
In the rare cases that data is shared in accordance with the data availability statement, it poses
a burden in terms of labor to lodge and respond to such requests, and if the data is large, may incur
an additional cost for hosting/transferring these large files.

@@

Some researchers would rather keep data for their private use.
Some may keep data private due to concerns over participant privacy.
This is a well-founded concern for identifiable data, however once identifying features are removed,
it is nearly impossible to re-identify genomic data @Venkatesaramani2021-tn unless the attackers have
pre-existing genomic data describing an individual or ancestry [@Schadt2012-md; @Gymrek2013-fz].
Therefore, sharing de-identified genomic data and metadata is likely to carry little risk to participant
privacy.
Still, much human genomic data and metadata resides in restricted databases due to the scope of
data sharing that was agreed to in the consent process.
For example, the consent form may only allow sharing of data to researchers investigating certain
diseases, which means a vetting process must be undertaken before sharing data with external
investigators.
In the interests of reuse and reproducibility, future studies could consider a consent form that allows
more broad sharing of deidentified data.

In order to maximize the value of shared data for reuse and reproducibility, it needs to be Findable,
Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq.
These principles are summarized in **Figure 5**.

@@

High throughput biology has already built a number of data repositories for specific data types
conforming to FAIR principles (e.g.: GEO, SRA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
However other domains lack well known specialized organized data repositories, such as evolutionary
biology, ecology, cell biology, social sciences and computing.
The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find a
repository that accepts data from these fields [@Pampel2013-ep].
If no specialized repository exists, then a general-purpose repository such as
[Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
should be considered.
Moreover, general recommendations have been proposed to enhance data FAIRness in ecology, but which
can also be applied to other disciplines [@Roche2015-aq].

@@

1. Depositing data to a specialized repository if possible, otherwise, a general-purpose repository.

@@

Once the quality of the data set has been established researchers may consider depositing it
to a data repository early, before publicization, as it has some advantages.
Most data repositories have an optional 12-month embargo period so researchers can share publicly
at a later date once they're happy with the overall project.
The data repository acts as an additional backup to the host institutions' own in case of data loss
or calamity.
By assuming the data and metadata can be sourced from a public location, the workflow developers ensure
that the code they share won't suffer from "file not found errors" during reproduction.
This ensures code and data are linked, which has been a priority for reproducibility [@Peng2011-ia],
and a source of problems for shared code [@Samuel2022-ha].

In addition to the experimental data, reference data that is critical to a project should be archived.
For example, in the field of genomics, gene function annotation sets are relied upon for a common
procedure called "pathway analysis" @Khatri2012-ye.
These annotation sets are regularly updated, but versions are rarely reported [@Wijesooriya2022-ly]
and finding historical data versions is sometimes difficult.
In order to guarantee future reproducibility, snapshots of those reference data should be archived
and shared if the license permits it.

@@

Documentation is the glue that binds a data science project together.
The manuscript/published article is the central artifact that outlines the research project and
links to the supporting materials.
From a reproducibility standpoint, the methods section is the most critical part.
It should be detailed enough so that other researchers can understand and replicate the
experiments/analysis and yield similar results.
Unfortunately, key details in bioinformatics data processing procedures are often omitted which limits
their reproducibility [@Simoneau2021-kp].
The "Materials Design Analysis Reporting" (MDAR) checklist for authors has been developed to assist in
comprehensive methodological reporting in the life sciences [@Mellor2022-up], and "Minimum Information
About a Bioinformatics Investigation" guidelines describe good reporting practices for computational
biology research [@Tan2010-is].

@@

As outlined above, one of the goals is to reduce the complexity of reproduction, and using a master
script is a simple way to do this.
It also makes the `README` documentation much simpler.
For example, the pathway analysis workflow we previously developed could be reproduced and inspected
with just six commands, starting from a new Ubuntu Linux install [@Ziemann2023-nm].
By minimizing the complexity of reproduction and writing the instructions clearly in the `README`,
reproducibility is made possible to a larger segment of the computational research community,
especially those who are not experts in the languages used.

@@

A lack of technical computer skills in many research groups is another culprit.
Specific workshop-based training in these areas could help, but the most scalable solution would be to
provide comprehensive beginner-friendly step-by-step guides for frequently used routines.
For example, our group has just released such a protocol that walks novice users through the process
of implementing functional enrichment analysis of gene expression data with the five pillars principles
@Ziemann2023-nm.
More such protocols are needed to make extreme reproducibility mainstream.
A compendium of publicly available learning materials around the five pillar concepts are
provided in the supplementary information.

@@

Reproducibility is an essential feature of reliable computational research yet current practices
are falling short of this goal.
Our recommendations for workflow-based research articles are summarized as:

@@

## Key points

* Irreproducibility of bioinformatics studies remains a significant problem.

* We present the *five pillars* framework, a set of best practices that enable extremely reproducible
workflows.

* Widespread adoption of these principles will enhance research reliability and speed translation of
basic research to tangible benefits.

@@

## Supplementary information

[Supplementary materials 1: A curated compendium of publicly available educational materials related
to five pillar concepts.](https://ziemann-lab.net/public/5pillars/practical_guides.html)

## Author biographies

**Mark Ziemann** is a Senior Lecturer in Biotechnology and Bioinformatics at Deakin University.
His research spans bioinformatics as applied to omics data analysis in biomedicine, as well as aspects
of research rigour.

**Samuel S. Minot** works as a Staff Scientist at Fred Hutchinson Cancer Center within the Vaccine and
Infectious Disease Division.
His research is focused on human microbiota associations with human health and disease.

**Pierre Poulain** is an Associate Professor at Université Paris Cité, Institut Jacques Monod, France.
Research directions include proteomics, bioinformatics and education/pedagogy.

**Anusuiya Bora** is a PhD student at Deakin University School of Life and Environmental Sciences.
Her research project involves improving the rigor of commonly used bioinformatics tools.
